---
title: "Bayesian capture-recapture inference with hidden Markov models"
description: "A two-day workshop with `R` and `Nimble`"
site: distill::distill_website
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

**What**: This is a workshop on Bayesian inference of animal demography. Hopefully, you will learn about how to infer demographic parameters (e.g. survival, dispersal). Our hope is to provide you with what you need to go your own path. The event is free of charge, and video-recorded. 

**For whom**: This is a workshop for ecologists. No previous experience with `Nimble` or Bayesian statistics is assumed, but knowledge of `R` is required. 

**How**: Through a combination of lectures and live demonstrations, you will get acquainted with the Bayesian approach, MCMC methods and the `R` `Nimble` package to fit single-site, multi-site, multi-state and multi-event models to capture-recapture data within the hidden Markov modeling (HMM) framework. 

**Who**: [O. Gimenez](https://twitter.com/oaggimenez), [C. R. Nater](https://twitter.com/chloe_nater), [S. Cubaynes](https://twitter.com/sarahcubaynes), [P. de Valpine](https://twitter.com/r_nimble), [M. Quéroué](https://twitter.com/maudqueroue)    

**When**: May 17-18, 2021     

**Where**: Online     

### Program

1. Welcoming words
2. Crash course on Bayesian statistics and MCMC algorithms
3. Free the modeler in you: Introduction to Nimble
4. What you see is not what you get: Hidden Markov models and capture-recapture data
5. Dead or alive: Survival estimation
6. On the move: Transition estimation
7. Known knowns, unknown knowns and unknowns: Uncertainty in state assignment
8. Skip your coffee break: Speed up MCMC convergence 
9. Take-home messages

### Requirements

* Install [`R`](https://cloud.r-project.org/) and [`RStudio`](https://rstudio.com/products/rstudio/download/#download)

* Install `Nimble` following [these guidelines](https://r-nimble.org/download).

* Install the following `R` packages: `tidyverse`, `MCMCvis`, `here`. You can install them all at once by running the following code in the `R` command line:

```{r eval = FALSE, echo = TRUE}
install.packages(c("tidyverse", "MCMCvis", "here"))
```


### Reuse {.appendix}

Text and figures are licensed under Creative Commons Attribution [CC BY 4.0](https://creativecommons.org/licenses/by/4.0/). Any computer code (R, HTML, CSS, etc.) in slides and worksheets, including in slide and worksheet sources, is also licensed under [MIT](https://github.com/oliviergimenez/bayesian-cr-workshop/blob/master/LICENSE.md). 

