---
title: "What you see is not what you get: Hidden Markov models and capture-recapture data"
author: "The team"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
header-includes: 
  - \usepackage{tikz}
  - \usepackage{pgfplots}
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")

library(tidyverse)
theme_set(theme_light())
update_geom_defaults("point", list(size = 2)) 
library(here)
library(nimble)
```

# Back to our survival example

--

+ We have $z$ survivors out of $n$ released animals with winter survival probability $\phi$

--

+ Our model so far:

\begin{align*}
   z &\sim \text{Binomial}(n, \phi) &\text{[likelihood]}
   \\
  \phi &\sim \text{Beta}(1, 1) &\text{[prior for }\phi \text{]} \\ 
\end{align*}

--

+ This is also:

\begin{align*}
   z_i &\sim \text{Bernoulli}(\phi), \; i = 1, \ldots, n &\text{[likelihood]}
   \\
  \phi &\sim \text{Beta}(1, 1) &\text{[prior for }\phi \text{]} \\ 
\end{align*}

--

+ Now what if we had several winters? Say five winters.

---
# Longitudinal data

+ $z_{i,t} = 1$ if individual $i$ alive at winter $t$, and $z_{i,t} = 2$ if dead.


.center.nogap[
```{r echo = FALSE}
nind <- 57
nocc <- 5
first <- rep(1, nind) # single cohort
z <- matrix(NA, nrow = nind, ncol = nocc)
phi <- 0.8
for (i in 1:nind){
  z[i,first[i]] <- 1
  for (t in (first[i]+1):nocc){
    z[i,t] <- rbinom(1, 1, phi * z[i,t-1]) # once you're dead z = 0, you remain dead
  }
}
z[z==0] <- 2 # 2 = dead, 1 = alive
colnames(z) <- paste0("winter ", 1:nocc)
z %>% 
  as_tibble() %>% 
  add_column(id = 1:nind, .before = "winter 1") %>%
  kableExtra::kable() %>%
  kableExtra::scroll_box(width = "100%", height = "400px")
#  kableExtra::kable_styling(font_size = 8, 
#                            latex_options = "scale_down")
```
]

???

+ Once you're dead, you remain dead.

---
# A model for longitudinal survival data

--

+ A model relies on assumptions.

--

+ The state of an animal at a given winter, alive or dead, is only dependent of its state the winter before. 

--

+ If an animal is alive in a given winter, the probability it survives to the next winter is $\phi$. 

--

+ The probability it dies is $1 - \phi$.

--

+ If an animal is dead a winter, it remains dead, unless you believe in zombies. 

--

+ This defines a Markov process.

---
# Markov process figure w/ tikz

---
# Transition matrix

+ The core of the Markov process is made of the transition probabilities.

+ For example, the probability of transitioning from state alive at $t-1$ to state alive at $t$ is $\Pr(z_t = 1 | z_{t-1} = 1) = \gamma_{1,1}$. It is the survival probability. 

+ The probability of dying over the interval $(t-1, t)$ is $\Pr(z_t = 2 | z_{t-1} = 1) = \gamma_{1,2}$.

+ Now if an animal is dead at $t-1$, then $\Pr(z_t = 1 | z_{t-1} = 2) = 0$ and $\Pr(z_t = 2 | z_{t-1} = 2) = 1$.

+ These probabilities can be packed in a transition matrix $\Gamma$:

\begin{align*}
\Gamma = 
\left(\begin{array}{cc} 
\gamma_{1,1} & \gamma_{1,2}\\ 
\gamma_{2,1} & \gamma_{2,2}
\end{array}\right) =
\left(\begin{array}{cc} 
\phi & 1 - \phi\\ 
0 & 1
\end{array}\right)
\end{align*}


---
# Markov process figure w/ tikz

Same figure w/ notation for $\gamma$ and $\phi$.

---
# Some Nimble code

---

<br><br>

This is the data we wish we had.

---
## Imperfect detection

+ Observation difficult
+ Imperfect detection
+ Add obs layer
+ Hidden markov models
+ Example with animal survival

---
## Implementation in Nimble

+ Some code

---
## A bit of history

+ probabilities à la main.

+ state-space which is HMM. 

---
# To-do list

+ Add decision trees, comme dans le workshop multievent

---
## Further reading

+ Zucchini et al. (2008, 2016) Hidden Markov Models for Time Series: An Introduction Using R

+ McClintock et al. (2020) Uncovering ecological state dynamics with hidden Markov models

+ Pradel 2005. 

+ Gimenez 2012. 