---
title: "Class 3 live demo: Intro to Nimble"
author: "The team"
date: "last updated: `r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")
library(tidyverse)
theme_set(theme_light(base_size = 14))
update_geom_defaults("point", list(size = 2)) 
library(here)
library(nimble)
```

## Introduction

In this demo, we show a simple example of how to use nimble. We use the survival example with $y = 19$ alive individuals out of $n = 57$ released. A binomial likelihood is assumed, with a uniform prior on survival. 

## Prepare everything 

First load the nimble package. 
```{r}
library(nimble)
```

Then define the model, likelihood and prior. 
```{r}
naive.survival.model <- nimbleCode({
  # prior
  phi ~ dunif(0, 1)
  # likelihood
  y ~ dbinom(phi, n)
})
```

Store the constants and data in lists.
```{r}
my.constants <- list(n = 57)
my.data <- list(y = 19)
```

Write a function to pick initial values for survival probability. 
```{r}
initial.values <- function() list(phi = runif(1,0,1))
```

Try the function.
```{r}
initial.values()
```

Specify the parameter you'd like to monitor, and for which you'd like to carry out posterior inference. 
```{r}
parameters.to.save <- c("phi")
```

Set the details of the MCMC run. Here we will use 5000 iterations, including 1000 for the burn-in period, and run two chains. 
```{r}
n.iter <- 5000
n.burnin <- 1000
n.chains <- 2
n.thin <- 1
```

## Run nimble

Run nimble. 
```{r, warning=FALSE, message=FALSE, eval = TRUE}
mcmc.output <- nimbleMCMC(code = naive.survival.model,     # model code  
                          data = my.data,                  # data
                          constants = my.constants,        # constants
                          inits = initial.values,          # initial values
                          monitors = parameters.to.save,   # parameters to monitor
                          thin = n.thin,                   # thinning interval (default = 1)
                          niter = n.iter,                  # nb iterations
                          nburnin = n.burnin,              # length of the burn-in
                          nchains = n.chains)              # nb of chains
```

## Post-process results

First, we have a look to the structure of the outputs. We have a list with two elements, each one is a chain with 4000 values (number of iterations minus the iterations we used for the burn-in) for survival. 
```{r}
str(mcmc.output)
```

Let's have a look to the first chain.
```{r}
head(mcmc.output$chain1)
```

We build a histogram of the values of the first chain, generated from the survival posterior distribution. 
```{r, echo = TRUE, fig.width = 7.5, fig.asp = 0.618, dev = "svg"}
mcmc.output %>%
  as_tibble() %>%
  janitor::clean_names() %>%
  ggplot() + 
  geom_histogram(aes(x = chain1[,"phi"]), color = "white") + 
  labs(x = "survival probability")
```

There are several packages that can take care of the post-processing for you. We will use [`MCMCvis`](https://joss.theoj.org/papers/10.21105/joss.00640) developed by Casey Youngflesh. 

Let's get the numerical summaries of the posterior distribution, along with the R-hat and the effective sample size. 
```{r}
library(MCMCvis)
MCMCsummary(mcmc.output, round = 2)
```

Now the trace plots and (a kernel density estimate of) the posterior distribution of survival. 
```{r, echo = TRUE, fig.width = 7.5, fig.asp = 0.618, dev = "svg"}
MCMCtrace(mcmc.output,
          pdf = FALSE) 
```

You may add the value of R-hat and the effective sample size on the graph. 
```{r, echo = TRUE, fig.width = 7.5, fig.asp = 0.618, dev = "svg"}
MCMCtrace(mcmc.output,
          pdf = FALSE,
          ind = TRUE,
          Rhat = TRUE,
          n.eff = TRUE) 
```

## Reproducibility for trouble-shooting and publishing
Being able to reproduce the exactly same initial values and MCMC chains can be very useful for pinpointing the source of errors, and also when publishing your code/workflows.

Here are three easy steps to make your workflow more reproducible: 

**1) Setting a defined initial seed at the start of your code**
```{r, echo = TRUE, eval = FALSE}
mySeed <- 0
set.seed(mySeed)
```

**2) Using "pre-sampled" initial values**
```{r, echo = TRUE, eval = FALSE}
initVals <- c(initial.values(), initial.values())
```

```{r, echo = TRUE, eval = FALSE}
mcmc.output <- nimbleMCMC(code = naive.survival.model,       
                          data = my.data,                  
                          constants = my.constants,        
                          inits = initVals,                # pre-sampled initial values
                          monitors = parameters.to.save,   
                          thin = n.thin,                   
                          niter = n.iter,                  
                          nburnin = n.burnin,              
                          nchains = n.chains)              

```

**3) Setting the MCMC seed**
```{r, echo = TRUE, eval = FALSE}
mcmc.output <- nimbleMCMC(code = naive.survival.model,       
                          data = my.data,                  
                          constants = my.constants,        
                          inits = initVals,                # pre-sampled initial values
                          monitors = parameters.to.save,   
                          thin = n.thin,                   
                          niter = n.iter,                  
                          nburnin = n.burnin,              
                          nchains = n.chains,
                          setSeed = mySeed)                # defined seed              

```


<!-- knitr::purl(here::here("worksheets","2_demo.Rmd"), documentation = 2) -->
